---
title: "`dplyr`-verbs, Piping & Function Composition"
output: 
  html_document:
    highlight: pygments
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```
```{r message=FALSE, warning=FALSE}
library(dplyr)
```


# Piping & Function composition

You have learnt about functions:

```{r}
my_vector <- runif(100) # 100 random values between 0 & 1

square <- function(x) x^2
deviation <- function(x) x - mean(x)
```

Imagine you want to compute the standard deviation of a vector from scratch. This is the square root of the mean squared deviation. We can write this as:

```{r}
sqrt(mean(square(deviation(my_vector))))
```

...but this is awkward, because to follow the sequence of computation, we have to read it from the inside out (first deviation, the square, then mean, then square root...). Alternatively, we can use intermediate objects:

```{r}
out <- deviation(my_vector)
out <- square(out)
out <- mean(out)
out <- sqrt(out)
out
```

But overwriting the same object is inefficient & looks stupid. So no-go! R has a third option: pipe operators `%>%` and `|>`. They allow you to chain functions together:

```{r}
my_vector |> 
  deviation() |> 
  square() |> 
  mean() |> 
  sqrt()
```

The easiest way to wrap your head around this is to pronounce the pipe in your head as "and then": take x, *and then* compute the deviation, *and then* square the result, *and then* take the mean, *and then* compute the square root.

It takes whatever is on the left-hand side, and passes it to the function on the right-hand side as the first argument, so `x |> f()` is the same as `f(x)`! We will talk about how to use it as another argument but the first later on.

The pipe allows you to read and write code sequentially, the same as you think!

# `dplyr`-verbs

`dplyr` is a grammar for data manipulation. It follows the logic of "verbs":

* `select()` allows you to pick variables by name
* `filter()` picks cases based on their values
* `mutate()` change existing or add new variables
* `arrange()` changes the ordering of rows
* `summarise()` reduces multiple values down to a single value

[Click here](https://dplyr.tidyverse.org/) to get to the `dplyr`-documentation, which is really comprehensive with many examples!

```{r}
library(gapminder)
data("gapminder")
```

Selecting:

```{r}
select(gapminder, country, year, lifeExp)
```

But for `dplyr`-verbs, the data is always the first argument, so we can pipe!

```{r}
gapminder |> select(country, year, lifeExp)
gapminder |> select(-c(continent, pop))

gapminder |> select(starts_with("c"))
```

Filtering:

```{r}
gapminder |> filter(continent == "Asia")
gapminder |> filter(year == 2007)

gapminder |> 
  filter(year == 2007 & continent %in% c("Asia", "Europe") & pop > 1e7)
```

Arranging:

```{r}
gapminder |> arrange(lifeExp)
gapminder |> arrange(desc(pop))
```

Summarising:

```{r}
gapminder |> 
  filter(year == 2007) |> 
  summarise(mean_life_exp = mean(lifeExp, na.rm = TRUE))
# Why not `mean()`? E.g. for grouped operations:

gapminder |> 
  group_by(continent) |> 
  summarise(mean_life_exp = mean(lifeExp, na.rm = TRUE)) |> 
  arrange(desc(mean_life_exp))
```

Mutating:

```{r}
gapminder |> mutate(pop_million = pop / 1e6)

gapminder |> 
  mutate(
    pop_million = pop / 1e6,
    gdp = gdpPercap * pop
  )
```

Other useful stuff:

```{r}
# count()

gapminder |> 
  filter(year == 2007) |> 
  count(continent) |> 
  arrange(desc(n))
```

```{r}
# n()

gapminder |> 
  filter(year == 2007) |> 
  group_by(continent) |> 
  summarise(N = n(), mean_life_exp = mean(lifeExp, na.rm = TRUE))
```

If there is one thing to keep in mind, it's the `dplyr`-verbs, because they make data manipulation & exploration really easy.

**Exercise:**

1. What countries were Top 10 in life expectancy (`lifeExp`) in 2007?
2. Compute overall by continent:
  a. Mean life expectancy
  b. mean GDP per capita
3. Add a new variable to the `gapminder` data set: Total GDP in billion USD (we can assume that GDP = `gdpPercap` $\times$ `pop`).

```{r}
# Your code here...
```