---
title: "Merging & manipulating data"
output: 
  html_document:
    highlight: pygments
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

```{r}
library(dplyr)
library(tidyr)
```

# Merging (joins)

# Reshaping (pivoting)

Notice anything with this data that is different than we have seen before?

```{r}
fertility <- readr::read_csv(here::here("data", "wbi_fertility.csv"))
fertility
```

**Wide** vs. **long** format. We have now encountered "wide" data for the first time. The `tidyr`-package comes with functions to reshape your data from wide into long and vice versa: `pivot_longer()` and `pivot_wider()`. Let's apply:

```{r}
fert_long <- 
  fertility |> 
  pivot_longer(
    cols = -starts_with("Country"),
    names_to = "year",
    values_to = "fertility"
  )

fert_long
```

To do the opposite, use `pivot_wider()`:

```{r}
fert_long |> 
  pivot_wider(
    names_from = "year",
    values_from = "fertility"
  )
```


# String manipulation & RegEx

```{r}
library(stringr)
```

(...)

This data set has a number of articles from the SPIEGEL's "Politics" & "Foreign" ressorts (from January 2023 to October 2023).

```{r}
spiegel <- readr::read_csv(here::here("data/spiegel.csv"))
```

Let's look for articles on Ukraine (mentioning Ukraine in the keywords):

```{r}
spiegel |> filter(str_detect(keywords, "Ukraine"))
```

Base R: grep*-functions ("global regular expression print"), which you may know from the terminal. 

* `grep()` returns the indices of matches
* `grepl()` returns TRUE or FALSE for every element

(There are also their weird cousins, like `gregexpr()` & `gregexec()`, which differ in what they return, but you will probably never need them)

```{r}
spiegel |> filter(grepl("Ukraine", keywords))
```

* Remove ticker articles.
* convert to lower, upper, title case.

```{r}
ticker_articles <- spiegel |> filter(str_detect(title, "News:"))
  
ticker_articles |> mutate(title = str_remove(title, "News: "))
```

For example, we can also use this to remove the base URL

```{r}
spiegel |> mutate(url = str_remove(url, "https://www.spiegel.de/")) |> select(url)
```

You can also replace strings:

```{r}
ticker_articles |> 
  mutate(title = str_replace(title, "News:", "LOOK!")) |> 
  select(title)
```

Or with base R:

```{r}
ticker_articles |> 
  mutate(title = gsub("News:", "Also Works!", title)) |> 
  select(title)
```

Converting to upper case, lower case, title case etc:

```{r}
title <- spiegel$title[1]

str_to_lower(title) # base: tolower()
str_to_upper(title) # base: toupper()
str_to_title(title)
str_to_sentence(title)
```

**Exercise:**

* Find all articles that mention Joe Biden in the title. *Bonus:* Find all articles that mention Joe Biden or Donald Trump. Replace both "Joe Biden" and "Donald Trump" with strings of your choice.

```{r}

```

* Find all articles on Ukraine. Which month had the most articles on Ukraine? Any idea why? Which month had the least? *Tip:* You can use `lubridate::month()` to extract the month from a date/datetime. Set `label = TRUE` to get the name of the month, not just its number.

```{r}

```

* Do the same for Israel. Which month had the most articles here?

```{r}

```
